\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2022-08-29 sections/methods.tex
Change: Edit the supplementary section for draft v2
Change: Update Results section for draft v3
Also, changes the figure order (code remains unchanged)

  \subsection*{16S rRNA sequencing datasets}
  \vspace{-5mm}
    This study utilized two kinds of 16S rRNA sequencing datasets: biological datasets and mock datasets.
    Biological datasets are collections of sequencing reads obtained from naturally occurring microbial community samples.
    The current analysis used stool samples from a fecal microbiome transplant study of autism~\cite{Kang2017} as biological datasets.
    The study was composed of multiple sequencing runs and those runs that contained paired end reads (run 2 (10M reads), run 3 (750K reads) and run 4 (16M reads)), were downloaded from Qiita~\cite{qiita} (study ID 10532) and used as input sequences for the \ac{micone} pipeline.
    Sequences from both control (212 samples including neurotypical and donors) and autistic (126 samples) patients were included in the analyses, but the groups were analyzed independently in the pipeline after the \ac{op} step.
    The mock community 16S datasets are sequencing data obtained for artificially assembled collections of species in known proportions.
    The mock datasets used for this study, obtained from mockrobiota~\cite{Bokulich2016}, are labelled mock4, mock12 and mock16.
    The mock4 community is composed of 21 bacterial strains.
    Two replicate samples from mock4 contain all species in equal abundances, and two additional replicate samples contain the same species in unequal abundances.
    The mock12 community is composed of 27 bacterial strains that include closely related taxa with some pairs having only one to two nucleotide difference from one another.
    The mock16 community is composed of 49 bacteria and 10 Archaea, all represented in equal amount.

@@

  The \ac{micone} pipeline provides both a Python API together with a command-line interface and only uses a single configuration file (nextflow.config) to contain the configuration parameters.
  The \ac{micone} Python API provides several biom (\ac{otu} table) and network related functions and methods, enabling detailed comparison of counts tables and inferred networks if desired.
  It also contains helper functions that can help in parsing taxonomies and communicate with the \ac{ncbi} taxonomy database to query taxonomy by name or taxonomic IDs.
  The configuration file along with the run file (main.nf) lists the inputs, output and the steps to be performed during runtime, along with the parameters to be used (if different from defaults) for the various steps.
  Since the entire pipeline run is stored in the form of a text file (the configuration file), subsequent runs are highly reproducible and changes can be easily tracked using version control.
  It makes use the nextflow workflow manager~\cite{Tommaso2015} under the hood, making it readily usable on the local machine, cluster or cloud with minimal configuration change.
  It also allows for automatic parallelization of all possible processes, both within and across samples.
  The pipeline is designed to be modular: each tool or method is organized into modules which can be easily modified or replaced.
  This modular architecture simplifies the process of adding new tools (refer to modules section in the \ac{micone} documentation).
  % TODO: Write a DockerFile for the package
  % In addition to the Python package, the entire pipeline has been containerized into a Docker~\cite{Merkel1994} image (\hl{dockerhub link}) for easy deployment and setup.
  The main components of the pipeline are detailed in the subsequent sections.

@@

  \subsection*{Taxonomy Assignment (TA)}
  \vspace{-5mm}
  This module deals with assigning taxonomies to either the representative sequences of the \ac{otu}s or the \ac{esv}s.
  In order to assign taxonomies to a particular sequence a taxonomy database and a query tool are necessary.
  The taxonomy database contains the collection of 16S sequences of microorganisms of interest and the query tool allows one to compare a sequence of interest to all the sequences in the database to identify the best matches.
  Finally, a consensus method is used to identify the most probable match from the list of best matches.
  The pipeline incorporates \ac{gg} 13\_8~\cite{DeSantis2006}, SILVA 138~\cite{Quast2012} and the \ac{ncbi} (16S RefSeq as of Oct 2021)~\cite{Sayers2009} databases for taxonomy assignment.
  These databases were downloaded and built using the RESCRIPt qiime2 plugin~\cite{iiRESCRIPtReproducibleSequence2021}.
  The Naive Bayes classifier and the \ac{ncbi} blast used as the query tools in this study were from the \ac{qiime2} package and the parameters used were the defaults of the package.
  The consensus algorithm used is the default method used by the classifiers in \ac{qiime2}.
  The list of tools used in this step, along with their modules and references are provided in Table~\ref{tab:micone_tools}.

@@

  \subsection*{Network Inference (NI)}
  \vspace{-5mm}
  This module deals with the inference of co-occurrence associations from the processed \ac{otu} or \ac{esv} counts matrix.
  The input count matrices are collapsed to the Genus level (or any other required taxonomy level) using the group module at the OP step.
  These collapsed matrices are used as input to the network inference methods to produce association matrices at the appropriate taxonomy level.
  These associations can be represented as a network, with nodes representing taxonomies of the microorganisms and edges representing the associations between them.
  A null model is created by re-sampling and bootstrapping the counts matrix and recalculating the correlations.
  These bootstrapped association matrices are used to calculate the significance of the inferred correlations by calculating the p-values against this null model~\cite{Watts2018}.
  The pipeline includes 4 methods for pairwise correlation metrics, and 6 methods for direct association metrics (refer Table~\ref{tab:micone_tools}).
  The Brown's p-value merging method~\cite{brown_400_1975} is used for combining p-values from the pairwise correlations methods to obtain a consensus p-value, which can be used to filter for significance.
  The bootstrapping and p-value calculations are only performed on the correlations-based methods.
  In the final module of this step, the consensus algorithms are used to create the final consensus network using associations from all the network inference methods (except Pearson and Spearman, by default).
  The output of this step are co-occurrence association networks encoded in the JSON format (refer Supplementary section) and which can also be exportable to a variety of network formats.
  The list of tools used in this step, along with their modules and references are provided in Table~\ref{tab:micone_tools}.

  \subsection*{Consensus Network and p-value merging}

@@

  \subsubsection*{Notation}
  \vspace{-5mm}
  This section defines notation for the network variability analysis performed for Figure~\ref{fig:figure2}.

@@

   The variance contributed by each step of the pipeline is calculated for every component in $\mathbf{C}$ matrix through ANOVA using the Python statsmodels~\cite{seaboldStatsmodelsEconometricStatistical2010} package and is shown in Figure~\ref{fig:figure2}A.
  The total variance for the network is calculated by adding the variances for each connection.
  The merged network table $\mathbf{\bar{N}}$ is used as the input to the PCA analysis to generate Figure~\ref{fig:figure2}B.

@@

  For each method, we generated the OTU table depicting the abundances of species and used this as input to generate association networks using \ac{micone} pipeline.
  The interaction matrix was used as the source of expected (true) interactions and the associations predicted using \ac{micone} were the source of predicted interactions.
  Finally, for each dataset we evaluated the precision and sensitivity of the associations predicted by the individual network inference methods as well as the consensus (Figure \ref{fig:figure6}).

@@

  \subsubsection*{DC step}
  \vspace{-5mm}
  In order to compare the representative sequences generated by the various methods in the DC step, we employ both the weighted~\cite{Lozupone2007} (Figure~\ref{fig:figure3}A) and unweighted UniFrac method~\cite{Lozupone2005} (Figure~\ref{fig:figure3}B).
  The UniFrac distance metric (unique fraction metric) is a beta-diversity measure that computes the distance between two sets of taxa as the fraction of the branch length of the tree that leads to descendants from either one environment or the other, but not both~\cite{Lozupone2005}
  The weighted UniFrac distance metric takes into account the abundances of the representative sequences when calculating shared and unshared branch lengths, whereas the unweighted UniFrac distance metric does not and hence gives equal weights to each sequence.
  In Figure~\ref{fig:figure3} the distances between methods are the average distance calculated between the reference sequence distribution for a pair of methods generated for every corresponding sample in the dataset.
  All UniFrac calculations were performed using the \texttt{scikit-bio}~\cite{thescikit-biodevelopmentteamScikitbioBioinformaticsLibrary2022} v0.5.6 Python package.

  \subsubsection*{TA step}
  \vspace{-5mm}
  In Figure~\ref{fig:figure4}C we use the Bray-Curtis distance metric to calculate the distance between the predicted (using the taxonomy databases in the TA step) and expected taxonomic distribution.
  The Bray-Curtis distance is used to quantify the compositional dissimilarity between two different taxonomic distributions.
  It is defined as:
  $$\sum{|u_i-v_i|} / \sum{|u_i+v_i|}$$
  The Bray-Curtis distance calculations were performed using the \texttt{scipy}~\cite{virtanenSciPyFundamentalAlgorithms2020} v1.8.0 Python package.

  \subsubsection*{NI step}
  \vspace{-5mm}
  In Figure~\ref{fig:figure6} we calculate the accuracy and robustness of the inferred association networks (using the various network inference algorithms and the consensus methods) against the original interaction used to create the taxonomic distribution.
  We use precision and sensitivity to calculate the accuracy of the predictions.

>>>>

  \subsection*{16S rRNA sequencing datasets}
  \vspace{-5mm}
    This study utilized two kinds of 16S rRNA sequencing datasets: biological datasets and mock datasets.
    Biological datasets are collections of sequencing reads obtained from naturally occurring microbial community samples.
    The current analysis used stool samples from a fecal microbiome transplant study of autism~\cite{Kang2017} as biological datasets.
    The study was composed of multiple sequencing runs and those runs that contained paired end reads (run 2 (10M reads), run 3 (750K reads) and run 4 (16M reads)), were downloaded from Qiita~\cite{qiita} (study ID 10532) and used as input sequences for the \ac{micone} pipeline.
    Sequences from both control (212 samples including neurotypical and donors) and autistic (126 samples) patients were included in the analyses, but the groups were analyzed independently in the pipeline after the \ac{op} step.
    All the network analyses in the study, unless explicitly mentioned, are performed indeed on the healthy and autistic samples in the fecal microbiome transplant study.
    However, all other analysis (preceding the NI step) are performed together on all the samples in the fecal microbiome transplant study.
    The mock community 16S datasets are sequencing data obtained for artificially assembled collections of species in known proportions.
    The mock datasets used for this study, obtained from mockrobiota~\cite{Bokulich2016}, are labelled mock4, mock12 and mock16.
    The mock4 community is composed of 21 bacterial strains.
    Two replicate samples from mock4 contain all species in equal abundances, and two additional replicate samples contain the same species in unequal abundances.
    The mock12 community is composed of 27 bacterial strains that include closely related taxa with some pairs having only one to two nucleotide difference from one another.
    The mock16 community is composed of 49 bacteria and 10 Archaea, all represented in equal amount.

@@

  The \ac{micone} pipeline provides both a Python API together with a command-line interface and only uses a single configuration file (nextflow.config) to contain the configuration parameters.
  The \ac{micone} Python API provides several biom (\ac{otu} table) and network related functions and methods, enabling detailed comparison of counts tables and inferred networks if desired.
  Exploring the effects of these combinations of methods on the resultant networks is difficult and inconvenient since different tools differ in their input and output formats and require interconversions between the various formats.
  The pipeline facilitates this comparative exploration by providing a variety of modules for interconversion between various formats, and by allowing for easy incorporation of new tools as modules.
  It also contains helper functions that can help in parsing taxonomies and communicate with the \ac{ncbi} taxonomy database to query taxonomy by name or taxonomic IDs.
  The configuration file along with the run file (main.nf) lists the inputs, output and the steps to be performed during runtime, along with the parameters to be used (if different from defaults) for the various steps.
  Since the entire pipeline run is stored in the form of a text file (the configuration file), subsequent runs are highly reproducible and changes can be easily tracked using version control.
  It makes use the nextflow workflow manager~\cite{Tommaso2015} under the hood, making it readily usable on the local machine, cluster or cloud with minimal configuration change.
  It also allows for automatic parallelization of all possible processes, both within and across samples.
  The pipeline is designed to be modular: each tool or method is organized into modules which can be easily modified or replaced.
  This modular architecture simplifies the process of adding new tools (refer to modules section in the \ac{micone} documentation).
  % TODO: Write a DockerFile for the package
  % In addition to the Python package, the entire pipeline has been containerized into a Docker~\cite{Merkel1994} image (\hl{dockerhub link}) for easy deployment and setup.
  The main components of the pipeline are detailed in the subsequent sections.

@@

  \subsection*{Taxonomy Assignment (TA)}
  \vspace{-5mm}
  This module deals with assigning taxonomies to either the representative sequences of the \ac{otu}s or the \ac{esv}s.
  In order to assign taxonomies to a particular sequence a taxonomy database and a query tool are necessary.
  The taxonomy database contains the collection of 16S sequences of microorganisms of interest and the query tool allows one to compare a sequence of interest to all the sequences in the database to identify the best matches.
  Finally, a consensus method is used to identify the most probable match from the list of best matches.
  The pipeline incorporates \ac{gg} 13\_8~\cite{DeSantis2006}, SILVA 138~\cite{Quast2012} and the \ac{ncbi} (16S RefSeq as of Oct 2021)~\cite{Sayers2009} databases for taxonomy assignment.
  SILVA and \ac{gg} are two popular 16S databases used for taxonomy identification and the \ac{ncbi} RefSeq nucleotide database contains 16S rRNA sequences as a part of two BioProjects - 33175 and 33317.
  The three databases vastly differ in terms of their last update status - \ac{gg} was last updated on May 2013, SILVA was last updated on August 2020 at the time of writing and \ac{ncbi} is updated regularly as new sequences are curated.
  These databases were downloaded and built using the RESCRIPt \ac{qiime2} plugin~\cite{iiRESCRIPtReproducibleSequence2021}.
  The Naive Bayes classifier and the \ac{ncbi} blast used as the query tools in this study were from the \ac{qiime2} package and the parameters used were the defaults of the package.
  The consensus algorithm used is the default method used by the classifiers in \ac{qiime2}.
  The assignment between SILVA and the other two databases were originally significantly different ($40\%$ mismatch) even at the Phylum level.
  However, this was fixed by making minor adjustments to the taxonomic names, such as changing Bacteroidota to Bacteroidetes in the SILVA Phylum assignments.
  The full list of changes can be found in \texttt{figure4ab_data.py} in the data and scripts  repository.
  The list of tools used in this step, along with their modules and references are provided in Table~\ref{tab:micone_tools}.

@@

  \subsection*{Network Inference (NI)}
  \vspace{-5mm}
  This module deals with the inference of co-occurrence associations from the processed \ac{otu} or \ac{esv} counts matrix.
  The input count matrices are collapsed to the Genus level (or any other required taxonomy level) using the group module at the OP step.
  These collapsed matrices are used as input to the network inference methods to produce association matrices at the appropriate taxonomy level.
  These associations can be represented as a network, with nodes representing taxonomies of the microorganisms and edges representing the associations between them.
  A threshold of 0.3 was set for the correlation-based methods (\ac{sparcc}, Propr, Spearman and Pearson) and a threshold of 0.01 was set for the other direct association methods.
  After filtering out edges based on the threshold, the isolated nodes are removed.

  A null model is created by re-sampling and bootstrapping the counts matrix and recalculating the correlations.
  These bootstrapped association matrices are used to calculate the significance of the inferred correlations by calculating the p-values against this null model~\cite{Watts2018}.
  The pipeline includes 4 methods for pairwise correlation metrics, and 6 methods for direct association metrics (refer Table~\ref{tab:micone_tools}).
  The Brown's p-value merging method~\cite{brown_400_1975} is used for combining p-values from the pairwise correlations methods to obtain a consensus p-value, which can be used to filter for significance.
  The bootstrapping and p-value calculations are only performed on the correlations-based methods.
  In the final module of this step, the consensus algorithms are used to create the final consensus network using associations from all the network inference methods (except Pearson and Spearman, by default).
  The output of this step are co-occurrence association networks encoded in the JSON format (refer Supplementary section) and which can also be exportable to a variety of network formats.
  The list of tools used in this step, along with their modules and references are provided in Table~\ref{tab:micone_tools}.

  \subsection*{Consensus network and p-value merging}

@@

  \subsubsection*{Notation}
  \vspace{-5mm}
  This section defines notation for the network variability analysis performed for Figure~\ref{fig:figure6}.

@@

   The variance contributed by each step of the pipeline is calculated for every component in $\mathbf{C}$ matrix through ANOVA using the Python statsmodels~\cite{seaboldStatsmodelsEconometricStatistical2010} package and is shown in Figure~\ref{fig:figure6}A.
  The total variance for the network is calculated by adding the variances for each connection.
  The merged network table $\mathbf{\bar{N}}$ is used as the input to the PCA analysis to generate Figure~\ref{fig:figure6}B.

@@

  For each method, we generated the OTU table depicting the abundances of species and used this as input to generate association networks using \ac{micone} pipeline.
  The interaction matrix was used as the source of expected (true) interactions and the associations predicted using \ac{micone} were the source of predicted interactions.
  Finally, for each dataset we evaluated the precision and sensitivity of the associations predicted by the individual network inference methods as well as the consensus (Figure \ref{fig:figure5}).

@@

  \subsubsection*{DC step}
  \vspace{-5mm}
  In order to compare the representative sequences generated by the various methods in the DC step, we employ both the weighted~\cite{Lozupone2007} (Figure~\ref{fig:figure2}A) and unweighted UniFrac method~\cite{Lozupone2005} (Figure~\ref{fig:figure2}B).
  The UniFrac distance metric (unique fraction metric) is a beta-diversity measure that computes the distance between two sets of taxa as the fraction of the branch length of the tree that leads to descendants from either one environment or the other, but not both~\cite{Lozupone2005}
  The weighted UniFrac distance metric takes into account the abundances of the representative sequences when calculating shared and unshared branch lengths, whereas the unweighted UniFrac distance metric does not and hence gives equal weights to each sequence.
  In Figure~\ref{fig:figure2} the distances between methods are the average distance calculated between the reference sequence distribution for a pair of methods generated for every corresponding sample in the dataset.
  All UniFrac calculations were performed using the \texttt{scikit-bio}~\cite{thescikit-biodevelopmentteamScikitbioBioinformaticsLibrary2022} v0.5.6 Python package.

  \subsubsection*{TA step}
  \vspace{-5mm}
  In Figure~\ref{fig:figure3}C we use the Bray-Curtis distance metric to calculate the distance between the predicted (using the taxonomy databases in the TA step) and expected taxonomic distribution.
  The Bray-Curtis distance is used to quantify the compositional dissimilarity between two different taxonomic distributions.
  It is defined as:
  $$\sum{|u_i-v_i|} / \sum{|u_i+v_i|}$$
  The Bray-Curtis distance calculations were performed using the \texttt{scipy}~\cite{virtanenSciPyFundamentalAlgorithms2020} v1.8.0 Python package.

  \subsubsection*{NI step}
  \vspace{-5mm}
  In Figure~\ref{fig:figure5} we calculate the accuracy and robustness of the inferred association networks (using the various network inference algorithms and the consensus methods) against the original interaction used to create the taxonomic distribution.
  We use precision and sensitivity to calculate the accuracy of the predictions.
